---
title: "CleaningMortality"
author: "Ainsley Lotito"
date: "2025-11-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE) 
library(readr)  
library(dplyr)
library(tidyverse)
library(forcats)
library(lubridate) 
library(naniar) 
library(plotly) 
library(ggplot2)
```


## Mortality Data

We will start by cleaning the mortality dataset from the national park service. 
Let's start by reading in the data and looking at the columns. 

```{r}
mortality <- read_csv("../data/NPS-Mortality-Data-CY2007-to-CY2024-Released-August-2024.csv", col_names = TRUE)
head(mortality) 
dim(mortality)
``` 

Mortality read in correctly, we have 9 variables and 4,635 observations. 

We need to reclassify some variables, rename them, and reclassify not reported information as NA. 
```{r} 
mortality <- mortality %>% 
  mutate(incident_date = mdy(`Incident Date`),
         year = year(incident_date)) %>% #creating year variable to aid with merging with other df
  rename(park = `Park Name`,
         death_cause = `Cause of Death`,
         death_group = 4,
         age_range = `Age Range`) %>% 
  select(-1,) 
colnames(mortality) <- tolower(colnames(mortality))
head(mortality)
``` 

We are going to explore missing data by reclassifying "Undetermined" and "Not reported" data in a new dataframe and look through some patterns to see if specific causes of death have higher proportions of missing data. 
```{r}
#changing "Not Reported" to be NA in different dataset
mortality_miss <- mortality %>% 
  mutate(across(c(intent, death_group, death_cause),
                ~ na_if(., "Undetermined")),
         across(c(age_range, activity),
                ~ na_if(., "Not Reported"))) 
gg_miss_var(mortality_miss)
gg_miss_upset(mortality_miss) 

mortality_miss %>% 
  mutate(missing = is.na(intent) & is.na(age_range) & is.na(activity)) %>% 
  group_by(death_cause, missing) %>% 
  count()

mortality_miss %>% 
  mutate(missing =  is.na(age_range)) %>% 
  group_by(death_cause, missing) %>% 
  count()

#add on later to summarize by proportions and which have highest mortality counts
```


We are going to subset our data to include years 2007-2016 to be consistent with the other dataset.  
```{r} 
#seeing how many obs we are cutting out

 mortality %>% 
  filter(year %in% c(2020:2024)) %>% 
   nrow()
#we are cutting out 2017 ()

mortality <- mortality %>% 
  filter(year %in% c(2007:2019))
dim(mortality) 
unique(mortality$year) 
```

We now have 3,262 observations (it removed 1,373 obsevations)  


# Exploratory analysis 

We are interested in looking at patterns of mortality in our data before we join the datasets to think through maps that would be interesting. 
First, we are going to plan out a chloropleth map that organizes mortality counts in the parks over the whole period (2007-2019) and by each year individually. 

We will start by analyzing this with the summarize function first to look at which parks have the highest mortality counts over each year. 
```{r}
#raw count
mortality %>% 
  group_by(park) %>%  
  summarize(mortality_count = n()) %>% 
  arrange(desc(mortality_count))  


#top 3 by year 
mortality %>% 
  group_by(park, year) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(year) %>% 
  slice_max(mortality_count, n = 3)

``` 

We also want to look into the causes of death and how those contribute to mortality counts. 
```{r}
#raw count
mortality %>% 
  group_by(death_group) %>%  
  summarize(mortality_count = n()) %>% 
  arrange(desc(mortality_count)) 

#top 3 by year 
mortality %>% 
  group_by(death_group, year) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(year) %>% 
  slice_max(mortality_count, n = 3) 

#let's do it again by age group 
#raw count
mortality %>% 
  group_by(death_group, age_range) %>%  
  summarize(mortality_count = n()) %>% 
  arrange(desc(mortality_count)) 

#top 3 by year 
mortality %>% 
  group_by(death_group, age_range, year) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(year) %>% 
  slice_max(mortality_count, n = 3)  

#top 3 by park 
mortality %>% 
  group_by(death_group, age_range, park) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(death_group,age_range) %>% 
  slice_max(mortality_count, n = 1) 
``` 

Let's explore some leaflet bar charts with different information on cause of death.  

```{r}
library(plotly)

#starting simple plot  
cod_count <- mortality %>% 
  group_by(death_group) %>%  
  summarize(mortality_count = n(), .groups = "drop")

cod_plot <-plot_ly(
  data = cod_count,
  x = ~death_group,
  y = ~mortality_count, 
  name = "Death Counts by Cause of Death", 
  type = "bar"
) 
cod_plot <- cod_plot %>% layout(yaxis = list(title = 'Mortality Count'), barmode = 'group') %>% layout(xaxis = list(title = 'Cause of Death'), barmode = 'group')

cod_plot
``` 
We can probably condense "Legal Intervention" in the Other category since there are only 2 deaths over the entire period. These deaths are due to people being killed or dying as a result of injuries from a law enforcement officer. 

We are going to condense the Medical categories together, since there is no information about whether the PA caused the medical event to occur, just that it happened simultaneously. 

I am going to create a new variable that has these cleaned up, and then summarize before creating a new basic plot then moving on with more details. 

```{r}
mortality <- mortality %>% 
  mutate(cod = case_when( 
    # Medical
    death_group %in% c("Medical - Not during physical activity", "Medical - Unknown", "Medical - During physical activity") ~ "Medical", 
    
    #Condensing Legal Intervention with Other
    death_group %in% c("Legal Intervention", "Other") ~ "Other", 
    
    #All other categories Consistent
    TRUE ~ death_group #keep original death_group codings
  ))
```

### Saving cleaned dataset 
We are going to save the cleaned mortality df to our data folder so it can be joined with the National Park dataframe for future analysis. 

```{r}
write.csv(mortality, "../data/mortality_clean.csv", row.names = TRUE)
```


Now, let's create two new plots with the new categorization: 
```{r} 
#COD
#starting simple plot  
cod_count <- mortality %>% 
  group_by(cod) %>%  
  summarize(mortality_count = n(), .groups = "drop")

cod_plot <-plot_ly(
  data = cod_count,
  x = ~cod,
  y = ~mortality_count, 
  name = "Death Counts by Cause of Death", 
  type = "bar"
) 
cod_plot <- cod_plot %>% layout(yaxis = list(title = 'Mortality Count'), barmode = 'group') %>% layout(xaxis = list(title = 'Cause of Death'), barmode = 'group')

cod_plot 

```  

Now, I want to explore how cause of death changes based on some other facets. 
I want to explore how the distribution of deaths by age group change depending on COD (facet by COD, bars are age groups).  
```{r cod_by_age, fig.height = 25, fig.width = 9}
 
#plotting 
p_age <- ggplot(cod_by_age, aes(x = age_range, y = mortality_count)) +
  geom_col() +
  facet_wrap(~ cod, ncol = 2, scales = "free_x") +
  labs(title = "Death Counts by Cause of Death and Age",
       x = "Victim Age Group", 
       y = "Death Count") + 
  theme_classic() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
    scale_y_continuous(breaks = c(seq(from = 0, to = 200, by = 25)
)) +
  theme(axis.text.x = element_text (size = 8),
        strip.text = element_text(size = 11, face = "bold"))

cod_by_age_plot <- ggplotly(p_age)
cod_by_age_plot  

#look into shiny embedding https://testing-apps.shinyapps.io/flexdashboard-shiny-embedding/

```  

```{r}
mortality <- read.csv("../data/mortality_clean.csv")


#shiny in here 
library(shiny)

# summarizing cod by age
cod_by_age <- mortality %>% 
  group_by(cod, age_range) %>%  
  summarize(mortality_count = n(), .groups = "drop")

cod_order <- cod_by_age %>%
  group_by(cod) %>%
  summarize(total_mort = sum(mortality_count)) %>%
  arrange(desc(total_mort))

cod_by_age$cod <- factor(cod_by_age$cod, levels = cod_order$cod)

#UI Function
ui <- fluidPage( #creates a fluidPage
  sidebarLayout( #creates sidebar on the left
    sidebarPanel( #what's in the sidebar
      selectInput("cod_choice", "Select Cause of Death:", 
                  choices = levels(cod_by_age$cod)) #cod_choic is input ID, "Select Cause of Death:" is label shown, choices -- populate the dropdown with faactor levels from the cod column
    ),
    mainPanel( #starting the main panel
      plotlyOutput("codPlot") #placeholder for a plotly plot
    )
  )
)

server <- function(input, output) { #server- takes inputs, produce output
  output$codPlot <- renderPlotly({ #create reacative plotly output named "codPlot". everything INSIDE CURLY runs when the user changes the dropdown
    
    filtered <- cod_by_age %>% 
      filter(cod == input$cod_choice) #only include rows where cod= user's selection
    
    p <- ggplot(filtered, aes(x = age_range, y = mortality_count)) +
      geom_col() +
      labs(
        title = paste("Deaths by Age for:", input$cod_choice),
        x = "Victim Age Group",
        y = "Death Count"
      ) +
      theme_classic()
    #ggplot of the filtered data 
    ggplotly(p) #turns into plotly
  })
}

shinyApp(ui, server) #ui=page layout for shiny, server= plotly 
#runs the App using the UI and server we just defined

#note: when setting up, make sure to include runtime:shiny???
#---
#title: "COD Dashboard"
#output: 
 # flexdashboard::flex_dashboard:
 #   orientation: columns
#runtime: shiny
#---

```



This is the plot option that we decided to put in our flexdashboard, so there is some extra code cleaning it up. I also am going to continue exploration below, but we didn't end up using any of the following maps. 



Now, I want to look at which Parks have the highest death count for each of these categories. For each COD, I am going to show the Parks with the top 5-10 death counts for each COD.  
```{r, fig.width = 8, fig.height = 7}
cod_park <- mortality %>% 
  group_by(cod, park, age_range) %>%  
  summarize(mortality_count = n(), .groups = "drop") %>% 
  group_by(cod, park) %>% 
  slice_max(mortality_count, n = 3) 

p <- ggplot(cod_park, aes(
  x = age_range,
  y = mortality_count,
  fill = park,
  text = paste("Park:", park,
               "<br>Age Range:", age_range,
               "<br>Count:", mortality_count)
)) +
  geom_col(position = "dodge") +
  facet_wrap(~cod) +
  labs(x = "Age Range", y = "Mortality Count") +
  theme_bw()

ggplotly(p, tooltip = "text")
```